---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "man/figures/README-",
  fig.width = 7,
  out.width = NULL
)
```

# chronos

<!-- badges: start -->
[![R-CMD-check](https://github.com/tesselle/chronos/workflows/R-CMD-check/badge.svg)](https://github.com/tesselle/chronos/actions)

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)

[![DOI JSS](https://img.shields.io/badge/JSS-10.18637/jss.v093.c01-brightgreen)](https://doi.org/10.18637/jss.v093.c01)
<!-- badges: end -->

**_chronos_ is a heavily modified fork of [_ArchaeoPhases_](https://github.com/ArchaeoStat/ArchaeoPhases) v1.5, see the changelog for details.**

## Overview

Statistical analysis of archaeological dates and groups of dates. **chronos** allows to post-process Markov Chain Monte Carlo (MCMC) simulations from [ChronoModel](https://chronomodel.com), [Oxcal](https://c14.arch.ox.ac.uk/oxcal.html) or [BCal](https://bcal.shef.ac.uk). This package provides functions for the study of rhythms of the long term from the posterior distribution of a series of dates (tempo and activity plot). It also allows the estimation and visualization of time ranges from the posterior distribution of groups of dates (e.g. duration, transition and hiatus between successive phases).

```{r citation, echo=FALSE, comment=''}
utils::citation("chronos")
```

## Installation

You can install the released version of **chronos** from [CRAN](https://CRAN.R-project.org) with:

```{r cran-installation, eval=FALSE}
install.packages("chronos")
```

And the development version from [GitHub](https://github.com/) with:

```{r gh-installation, eval=FALSE}
# install.packages("remotes")
remotes::install_github("tesselle/chronos")
```

## Usage

These examples use data available through the [**fasti**](https://packages.tesselle.org/fasti/) package which is available in a separate [repository](https://tesselle.r-universe.dev). **fasti** provides MCMC outputs from ChronoModel, OxCal and BCal.

```{r fasti, eval=FALSE}
## Install the latest version
install.packages("fasti", repos = "https://tesselle.r-universe.dev")
```

```{r packages-load}
## Load packages
library(chronos)
```

Import a CSV file containing a sample from the posterior distribution:

```{r read-chronomodel}
## Read output from ChronoModel
path <- "chronomodel/ksarakil/"

## Events
path_events <- system.file(path, "Chain_all_Events.csv", package = "fasti")
(chrono_events <- read_chronomodel_events(path_events))

## Phases
path_phases <- system.file(path, "Chain_all_Phases.csv", package = "fasti")
(chrono_phases <- read_chronomodel_phases(path_phases))
```

**chronos** uses [**ggplot2**](https://github.com/tidyverse/ggplot2) for plotting information. This makes it easy to customize diagrams (e.g. using themes and scales).

### Analysis of a series of dates

```{r events-plot, fig.width=5, fig.align='default', fig.height=5, out.width='50%', fig.show='hold'}
## Plot the first event
plot(chrono_events, select = 1, interval = "hpdi")

## Plot all events
plot(chrono_events)
```

```{r tempo-plot, fig.width=5, fig.align='default', fig.height=5, out.width='50%', fig.show='hold'}
## Tempo plot
tp <- tempo(chrono_events, level = 0.95)
plot(tp)

## Activity plot
ac <- activity(chrono_events)
plot(ac)
```

### Analysis of a group of dates (phase)

```{r phases-boundaries}
boundaries(chrono_phases, level = 0.95)
```

```{r phases-plot, fig.height=3.5}
plot(chrono_phases)
```

```{r phases-order, fig.height=3.5}
## Set chronological order
## (from the oldest to the youngest phase)
set_order(chrono_phases) <- c("IUP", "Ahmarian", "UP", "EPI")
get_order(chrono_phases)
```

```{r succession-plot, fig.height=2}
plot(chrono_phases, level = 0.95)
```
