---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = '100%'
)
```

# ArchaeoPhases

<!-- badges: start -->
[![R-CMD-check](https://github.com/nfrerebeau/ArchaeoPhases/workflows/R-CMD-check/badge.svg)](https://github.com/nfrerebeau/ArchaeoPhases/actions)

[![CRAN Version](http://www.r-pkg.org/badges/version/ArchaeoPhases)](https://cran.r-project.org/package=ArchaeoPhases)
[![CRAN checks](https://cranchecks.info/badges/worst/ArchaeoPhases)](https://cran.r-project.org/web/checks/check_results_ArchaeoPhases.html)
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/ArchaeoPhases)](https://cran.r-project.org/package=ArchaeoPhases)

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)

[![DOI JSS](https://img.shields.io/badge/JSS-10.18637/jss.v093.c01-brightgreen)](https://doi.org/10.18637/jss.v093.c01)
<!-- badges: end -->

**This repository contains a heavily modified fork of the _ArchaeoPhases_ package. You might be interested in the [original package](https://github.com/ArchaeoStat/ArchaeoPhases).**

## Overview

Statistical analysis of archaeological dates and groups of dates. **ArchaeoPhases** allows to post-process Markov Chain Monte Carlo (MCMC) simulations from [ChronoModel](https://chronomodel.com), [Oxcal](https://c14.arch.ox.ac.uk/oxcal.html) or [BCal](https://bcal.shef.ac.uk). This package provides functions for the study of rhythms of the long term from the posterior distribution of a series of dates (tempo and activity plot). It also allows the estimation and visualization of time ranges from the posterior distribution of groups of dates (e.g. duration, transition and hiatus between successive phases).

```{r citation, echo=FALSE, comment=''}
utils::citation("ArchaeoPhases")
```

## Installation

You can install the released version of **ArchaeoPhases** from [CRAN](https://CRAN.R-project.org) with:

```{r cran-installation, eval=FALSE}
install.packages("ArchaeoPhases")
```

And the development version from [GitHub](https://github.com/) with:

```{r gh-installation, eval=FALSE}
# install.packages("remotes")
remotes::install_github("nfrerebeau/ArchaeoPhases")
```

## Usage

```{r packages-load}
## Load packages
library(ArchaeoPhases)
```

Import a CSV file containing a sample from the posterior distribution:

```{r read-chronomodel}
## Read output from ChronoModel
output_zip <- system.file("extdata/chronomodel.zip", package = "ArchaeoPhases")
output_csv <- utils::unzip(output_zip, exdir = tempdir())

(chrono_events <- read_chronomodel(output_csv[[1]], phases = FALSE))
(chrono_phases <- read_chronomodel(output_csv[[2]], phases = TRUE))
```

**ArchaeoPhases** uses [**ggplot2**](https://github.com/tidyverse/ggplot2) for plotting information. This makes it easy to customize diagrams (e.g. using themes and scales).

### Analysis of a series of dates

```{r events-plot, fig.width=7, fig.height=5}
plot(chrono_events) +
  ggplot2::theme_bw()
```

```{r tempo-plot, fig.width=5, fig.height=5, out.width='50%', fig.show='hold'}
## Tempo plot
tp <- tempo(chrono_events, level = 0.95)
plot(tp)

## Activity plot
ac <- activity(chrono_events)
plot(ac)
```

### Analysis of a group of dates (phase)

```{r phases-boundaries}
boundaries(chrono_phases, level = 0.95)
```

```{r phases-plot, fig.width=7, fig.height=3.5}
plot(chrono_phases)
```

```{r phases-order, fig.width=7, fig.height=3.5}
## Set chronological order
## (from the oldest to the youngest phase)
set_order(chrono_phases) <- c("IUP", "Ahmarian", "UP", "EPI")
get_order(chrono_phases)
```

```{r succession-plot, fig.width=7, fig.height=2}
plot(chrono_phases, level = 0.95)
```
