---
output: github_document
bibliography: vignettes/bibliography.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "man/figures/README-",
  fig.width = 7,
  out.width = NULL
)
```

# ArchaeoPhases

<!-- badges: start -->
[![R-CMD-check](https://github.com/ArchaeoStat/ArchaeoPhases/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ArchaeoStat/ArchaeoPhases/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/ArchaeoStat/ArchaeoPhases/branch/master/graph/badge.svg)](https://app.codecov.io/gh/ArchaeoStat/ArchaeoPhases)
[![Dependencies](https://tinyverse.netlify.com/badge/ArchaeoPhases)](https://cran.r-project.org/package=ArchaeoPhases)

[![r-universe](https://ArchaeoStat.r-universe.dev/badges/ArchaeoPhases)](https://ArchaeoStat.r-universe.dev){.pkgdown-devel}
[![CRAN Version](http://www.r-pkg.org/badges/version/ArchaeoPhases)](https://cran.r-project.org/package=ArchaeoPhases){.pkgdown-release}
[![CRAN checks](https://badges.cranchecks.info/worst/ArchaeoPhases.svg)](https://cran.r-project.org/web/checks/check_results_ArchaeoPhases.html){.pkgdown-release}
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/ArchaeoPhases)](https://cran.r-project.org/package=ArchaeoPhases){.pkgdown-release}

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)

[![DOI JSS](https://img.shields.io/badge/JSS-10.18637/jss.v093.c01-brightgreen)](https://doi.org/10.18637/jss.v093.c01)
<!-- badges: end -->

## Overview

Statistical analysis of archaeological dates and groups of dates. **ArchaeoPhases** allows to post-process Markov Chain Monte Carlo (MCMC) simulations from [ChronoModel](https://chronomodel.com) [@lanos2020], [Oxcal](https://c14.arch.ox.ac.uk/oxcal.html) [@bronkramsey2009] or [BCal](https://bcal.shef.ac.uk) [@buck1999]. This package provides functions for the study of rhythms of the long term from the posterior distribution of a series of dates (tempo and activity plot). It also allows the estimation and visualization of time ranges from the posterior distribution of groups of dates (e.g. duration, transition and hiatus between successive phases).

**Version 2.0 of ArchaeoPhases introduces a complete rewrite of the package: all functions have been renamed (see `news(Version >= "2.0", package = "ArchaeoPhases")`).**

```{r citation, echo=FALSE, comment=''}
utils::citation("ArchaeoPhases")
```

## Installation

You can install the released version of **ArchaeoPhases** from [CRAN](https://CRAN.R-project.org) with:

```{r cran-installation, eval=FALSE}
install.packages("ArchaeoPhases")
```

And the development version from [GitHub](https://github.com/) with:

```{r gh-installation, eval=FALSE}
# install.packages("remotes")
remotes::install_github("ArchaeoStat/ArchaeoPhases")
```

You can install the 1.x releases from the CRAN archives:

```{r archive-installation, eval=FALSE}
# install.packages("devtools")
devtools::install_version("ArchaeoPhases", version = "1.8")
```

## Usage

These examples use data available through the [**fasti**](https://packages.tesselle.org/fasti/) package which is available in a [separate repository](https://tesselle.r-universe.dev). **fasti** provides MCMC outputs from ChronoModel, OxCal and BCal.

```{r fasti, eval=FALSE}
## Install the latest version
install.packages("fasti", repos = "https://tesselle.r-universe.dev")
```

```{r packages-load}
## Load package
library(ArchaeoPhases)
```

Import a CSV file containing a sample from the posterior distribution:

```{r read-chronomodel}
## Read output from ChronoModel
path <- "chronomodel/ksarakil/"

## Events
path_events <- system.file(path, "Chain_all_Events.csv", package = "fasti")
(chrono_events <- read_chronomodel_events(path_events))

## Phases
path_phases <- system.file(path, "Chain_all_Phases.csv", package = "fasti")
(chrono_phases <- read_chronomodel_phases(path_phases))
```

### Analysis of a series of dates

```{r events-plot, fig.width=5, fig.align='default', fig.height=5, out.width='50%', fig.show='hold'}
## Plot the first event
plot(chrono_events[[1]], interval = "hpdi")

## Plot all events
plot(chrono_events)
```

```{r tempo-plot, fig.width=5, fig.align='default', fig.height=5, out.width='50%', fig.show='hold'}
## Tempo plot
tp <- tempo(chrono_events, level = 0.95)
plot(tp)

## Activity plot
ac <- activity(chrono_events)
plot(ac)
```

### Analysis of a group of dates (phase)

```{r phases-boundaries}
boundaries(chrono_phases, level = 0.95)
```

```{r phases-plot, fig.height=3.5}
plot(chrono_phases)
```

```{r succession-plot, fig.height=3.5}
plot(chrono_phases[, c("UP", "EPI"), ], range = "hiatus")
plot(chrono_phases[, c("UP", "EPI"), ], range = "transition")
```

## References
