---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = NULL
)
```

# ArchaeoPhases

<!-- badges: start -->
<!-- badges: end -->

**This repository contains a heavily modified fork of the _ArchaeoPhases_ package. You might be interested in the [original package](https://github.com/ArchaeoStat/ArchaeoPhases).**

## Overview

Statistical analysis of archaeological dates and groups of dates. **ArchaeoPhases** allows to post-process Markov Chain Monte Carlo (MCMC) simulations from [ChronoModel](https://chronomodel.com), [Oxcal](https://c14.arch.ox.ac.uk/oxcal.html) or [BCal](https://bcal.shef.ac.uk). It provides functions for the estimation and visualization of time ranges from the posterior distribution of a series of dates (e.g. transition and hiatus between successive phases).

To cite **ArchaeoPhases** in publications please use:

> Philippe, A. & Vibet, M.-A. (2020). Analysis of Archaeological Phases Using the R Package ArchaeoPhases. *Journal of Statistical Software, Code Snippets*, 93(1), 1-25. DOI [10.18637/jss.v093.c01](https://doi.org/10.18637/jss.v093.c01).

## Installation

You can install the released version of **ArchaeoPhases** from [CRAN](https://CRAN.R-project.org) with:

```{r cran-installation, eval=FALSE}
install.packages("ArchaeoPhases")
```

And the development version from [GitHub](https://github.com/) with:

```{r gh-installation, eval=FALSE}
# install.packages("remotes")
remotes::install_github("ArchaeoStat/ArchaeoPhases")
```

## Usage

```{r packages-load}
## Load packages
library(ArchaeoPhases)

library(ggplot2)
library(magrittr)
```

Import a CSV file containing a sample from the posterior distribution:

```{r read-chronomodel}
## Read output from ChronoModel
path_zip <- system.file("extdata/chronomodel.zip", package = "ArchaeoPhases")
path_csv <- utils::unzip(path_zip, exdir = tempdir())

chrono_events <- read_chronomodel(path_csv[[1]])
chrono_phases <- read_chronomodel(path_csv[[2]])
```

**ArchaeoPhases** uses [**ggplot2**](https://github.com/tidyverse/ggplot2) for plotting information. This makes it easy to customize diagrams (e.g. using themes and scales).

### Analysis of a series of dates

```{r events-plot, fig.width=7, fig.height=5}
plot(chrono_events) +
  ggplot2::theme_bw()
```

```{r tempo-plot, fig.width=3.5, fig.height=3.5, fig.show='hold'}
## Tempo plot
tp <- tempo(chrono_events, level = 0.95)
plot(tp) +
  ggplot2::theme_bw() +
  ggplot2::theme(legend.position = "bottom")

## Activity plot
ac <- activity(chrono_events)
plot(ac) +
  ggplot2::theme_bw()
```

### Analysis of a series of phases

```{r phases-plot, fig.width=7, fig.height=3.5}
## Get phases
phases <- as_phases(chrono_phases)
names(phases) <- c("EPI", "UP", "Ahmarian", "IUP")

plot(phases) +
  ggplot2::theme_bw()
```

```{r, fig.width=7, fig.height=3.5}
## Set chronological order
set_order(phases) <- c("IUP", "Ahmarian", "UP", "EPI")
get_order(phases)
```

```{r succession-plot, fig.width=7, fig.height=3.5}
plot(phases, level = 0.95) +
  ggplot2::theme_bw()
```
