---
output: github_document
bibliography: vignettes/bibliography.bib
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  fig.path = "man/figures/README-",
  fig.width = 7,
  out.width = NULL
)
```

# ArchaeoPhases

<!-- badges: start -->
[![R-CMD-check](https://github.com/ArchaeoStat/ArchaeoPhases/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/ArchaeoStat/ArchaeoPhases/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/ArchaeoStat/ArchaeoPhases/branch/master/graph/badge.svg?token=RVSFB6UI4T)](https://app.codecov.io/gh/ArchaeoStat/ArchaeoPhases)
[![Dependencies](https://tinyverse.netlify.com/badge/ArchaeoPhases)](https://cran.r-project.org/package=ArchaeoPhases)

[![r-universe](https://ArchaeoStat.r-universe.dev/badges/ArchaeoPhases)](https://ArchaeoStat.r-universe.dev){.pkgdown-devel}
[![CRAN Version](http://www.r-pkg.org/badges/version/ArchaeoPhases)](https://cran.r-project.org/package=ArchaeoPhases){.pkgdown-release}
[![CRAN checks](https://badges.cranchecks.info/worst/ArchaeoPhases.svg)](https://cran.r-project.org/web/checks/check_results_ArchaeoPhases.html){.pkgdown-release}
[![CRAN Downloads](http://cranlogs.r-pkg.org/badges/ArchaeoPhases)](https://cran.r-project.org/package=ArchaeoPhases){.pkgdown-release}

[![Project Status: Active – The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)

[![DOI JSS](https://img.shields.io/badge/JSS-10.18637/jss.v093.c01-brightgreen)](https://doi.org/10.18637/jss.v093.c01)
<!-- badges: end -->

## Overview

Statistical analysis of archaeological dates and groups of dates. **ArchaeoPhases** allows to post-process Markov Chain Monte Carlo (MCMC) simulations from [ChronoModel](https://chronomodel.com) [@lanos2020], [Oxcal](https://c14.arch.ox.ac.uk/oxcal.html) [@bronkramsey2009] or [BCal](https://bcal.shef.ac.uk) [@buck1999]. This package provides functions for the study of rhythms of the long term from the posterior distribution of a series of dates (tempo and activity plot). It also allows the estimation and visualization of time ranges from the posterior distribution of groups of dates (e.g. duration, transition and hiatus between successive phases).

**ArchaeoPhases v2.0 brings a comprehensive package rewrite, resulting in the renaming of nearly all functions. For more information, please refer to `news(Version >= "2.0", package = "ArchaeoPhases")`.**

```{r citation, echo=FALSE, comment=''}
cite <- utils::citation("ArchaeoPhases")
print(cite, bibtex = FALSE)
```

## Installation

You can install the released version of **ArchaeoPhases** from [CRAN](https://CRAN.R-project.org) with:

```{r cran-installation, eval=FALSE}
install.packages("ArchaeoPhases")
```

And the development version from [GitHub](https://github.com/) with:

```{r gh-installation, eval=FALSE}
# install.packages("remotes")
remotes::install_github("ArchaeoStat/ArchaeoPhases")
```

You can install the 1.x releases from the CRAN archives:

```{r archive-installation, eval=FALSE}
# install.packages("remotes")
remotes::install_version("ArchaeoPhases", version = "1.8")
```

## Usage

**ArchaeoPhases** v2.0 uses **aion** for internal date representation. Look at `vignette("aion")` before you start.

These examples use data available through the [**ArchaeoData**](https://github.com/ArchaeoStat/ArchaeoData) package which is available in a [separate repository](https://archaeostat.r-universe.dev). **ArchaeoData** provides MCMC outputs from ChronoModel, OxCal and BCal.

```{r fasti, eval=FALSE}
## Install data package
install.packages("ArchaeoData", repos = "https://archaeostat.r-universe.dev")
```

```{r packages-load, message=FALSE}
## Load package
library(ArchaeoPhases)
```

Import a CSV file containing a sample from the posterior distribution:

```{r read-chronomodel}
## Read output from ChronoModel
path <- "chronomodel/ksarakil/"

## Events
path_event <- system.file(path, "Chain_all_Events.csv", package = "ArchaeoData")
(chrono_events <- read_chronomodel_events(path_event))

## Phases
path_phase <- system.file(path, "Chain_all_Phases.csv", package = "ArchaeoData")
(chrono_phases <- read_chronomodel_phases(path_phase))
```

### Analysis of a series of dates

```{r events-plot, fig.width=5, fig.align='default', fig.height=5, out.width='50%', fig.show='hold'}
## Plot the first event
plot(chrono_events[, 1], interval = "hdr")

## Plot all events
plot(chrono_events)
```

```{r tempo-plot, fig.width=5, fig.align='default', fig.height=5, out.width='50%', fig.show='hold'}
## Tempo plot
tp <- tempo(chrono_events, level = 0.95)
plot(tp)

## Activity plot
ac <- activity(chrono_events)
plot(ac)
```

### Analysis of a group of dates (phase)

```{r phases-boundaries}
bound <- boundaries(chrono_phases, level = 0.95)
as.data.frame(bound)
```

```{r phases-plot}
## Plot all phases
plot(chrono_phases)
```

```{r succession-plot}
plot(chrono_phases[, c("UP", "EPI"), ], succession = "hiatus")
plot(chrono_phases[, c("UP", "EPI"), ], succession = "transition")
```

## References
